---
title: "Overview of Scry Methods"
author: "Will Townes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  BiocStyle::html_document:
    toc: false
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Overview of Scry Methods}
  %\usepackage[UTF-8]{inputenc}
---

# Basic Workflow

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
library(DuoClustering2018)
require(scry)
```

We illustrate the basic functionality of scry using a synthetic mixture of four
known cell types from the DuoClustering2018 package.

```{r}
sce<-sce_full_Zhengmix4eq()
#m<-counts(sce) #UMI counts
#cm<-as.data.frame(colData(sce))
```

## Feature Selection with Deviance

First we will rank genes based on deviance, to help identify the most
biologically informative genes. The actual deviance values are stored in the
rowData of the SingleCellExperiment object.

```{r fig.width=6, fig.height=4}
sce<-devianceFeatureSelection(sce, assay="counts", sorted=TRUE)
plot(rowData(sce)$binomial_deviance, type="l", xlab="ranked genes",
     ylab="binomial deviance", main="Feature Selection with Deviance")
abline(v=2000, lty=2, col="red")
```

We can see that the deviance drops sharply after about 2,000 genes. The
remaining genes are probably not informative so we discard them to speed up
downstream analysis.

```{r}
sce2<-sce[1:1000, ]
```

## Dimension Reduction with GLM-PCA

GLM-PCA can reduce the dimensionality of UMI counts to facilitate visualization
and/or clustering without needing any normalization.

```{r fig.width=6, fig.height=4}
set.seed(101)
sce2<-GLMPCA(sce2, 2, assay="counts")
fit<-metadata(sce2)$glmpca
pd<-cbind(as.data.frame(colData(sce2)), fit$factors)
ggplot(pd, aes(x=dim1, y=dim2, colour=phenoid)) + geom_point(size=.8) +
  ggtitle("GLM-PCA applied to high deviance genes")
```

The separation between B-cells, monocytes, and T-cells is clear. The separation
between naive cytotoxic and regulatory T-cells is less clear. Increasing the
number of latent factors from 2 to 10 can improve resolution of biological
clusters, but at the cost of slower computation.

## Dimension Reduction with Null Residuals

GLM-PCA can be slow for large datasets. A fast approximation is to fit a null
model of constant expression for each gene across cells, then fit standard
PCA to either the Pearson or deviance residuals from the null model.

```{r fig.width=6, fig.height=8}
sce<-nullResiduals(sce, assay="counts", type="deviance")
sce<-nullResiduals(sce, assay="counts", type="pearson")
sce2<-sce[1:1000, ] #use only the high deviance genes
pca<-function(Y, L=2, center=TRUE, scale=TRUE){
    #assumes features=rows, observations=cols
    res<-prcomp(as.matrix(t(Y)), center=center, scale.=scale, rank.=L)
    factors<-as.data.frame(res$x)
    colnames(factors)<-paste0("dim", 1:L)
    factors
}
pca_d<-pca(assay(sce2, "binomial_deviance_residuals"))
pca_d$resid_type<-"deviance_residuals"
pca_p<-pca(assay(sce2, "binomial_pearson_residuals"))
pca_p$resid_type<-"pearson_residuals"
cm<-as.data.frame(colData(sce2))
pd<-rbind(cbind(cm, pca_d), cbind(cm, pca_p))
ggplot(pd, aes(x=dim1, y=dim2, colour=phenoid)) + geom_point() +
  facet_wrap(~resid_type, scales="free", nrow=2) +
  ggtitle("PCA applied to null residuals of high deviance genes")
```

The null residuals approach still captures most of the biological structure,
but the resolution between clusters is diminished and there is more noise.

### Working with HDF5 matrices

For large datasets, it may be advantageous to keep the original data in a on-disk HDF5 file and avoid loading the full data in memory.

One can run `nullResiduals` on `HDF5Matrix`, `DelayedArray` matrices, and sparse matrices from the `Matrix` package with the same syntax used for the base matrix case.

We illustrate this with one dataset from the `TENxPBMCData` package.

```{r, eval=FALSE}
library(TENxPBMCData)
pbmc <- TENxPBMCData("pbmc4k")

pbmc <- nullResiduals(pbmc, assay="counts", type="deviance")

```
